A Formal Theory of Bidirectional Typing for Simple Types

### 2023-05-26 tex/BiSig.tex
[ tex ] Attempt to fix bidirectional pre-typed terms

\usepackage{bm}
\usepackage{braket}
\usepackage{mathtools}
\usepackage{mathpartir}
\usepackage[all,cmtip]{xy}
\usepackage{bussproofs}
\EnableBpAbbreviations

\usepackage{xfrac,xspace}
\usepackage{xcolor}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage[capitalise,noabbrev]{cleveref}
\usepackage[inline]{enumitem} % for environment enumerate*

@@

\section{Bidirectional simply typed \texorpdfstring{$\lambda$}{lambda}-calculus}\label{sec:bi-stlc}
\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta \mid \Gamma \vdash \isTerm{t} \Rightarrow A}
    \quad\text{and}\quad \boxed{\Theta \mid \Gamma \vdash \isTerm{t} \Leftarrow A}
    \quad\text{where $\Theta \vdash \Gamma$ and $\Theta \vdash A$} 
    \\
    \inferrule{(x : A) \in \Gamma}{\Theta \mid \Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
    \and
    \inferrule{\Theta \mid \Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Theta \mid \Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
    \and
    \inferrule{\Theta \mid \Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash \isTerm{t\uparrow} \Leftarrow A}\;(\text{sub})
    \and
    \inferrule{\Theta \mid \Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Theta \mid \Gamma \vdash \isTerm{u} \Leftarrow A}{\Theta \mid \Gamma \vdash \isTerm{t\;u} \Rightarrow B}
    \and
    \inferrule{\Theta \mid \Gamma, \isTerm{x} : A \vdash \isTerm{t} \Rightarrow B}{\Theta \mid \Gamma \vdash \isTerm{\lam{x}t} \Rightarrow A \bto B}
  \end{mathpar}
  \caption{The extrinsic style of bidirectional simply typed $\lambda$-calculus with type variables}
  \label{fig:bi-stlc}
\end{figure}
\begin{remark}
  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
\end{remark}

@@

\subsection{Binding Signatures}
\begin{definition}\label{def:binding-signature}
  For a type signature $\Sigma$, a \emph{binding signature} $\Omega$ consists of a set $O$ and a function
  \[
    \mathit{ar}\colon O \to \sum_{\Xi : \UU} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
  \]
  Each inhabitant $o: O$ is meant to represent a term construct $\termOp_o$ in a simple type theory with a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A\right)$
  instead of a number as its arity $\arity_o$ where
  \begin{enumerate}
    \item $A : \Type_\Sigma(\Xi)$ is the target type of $\termOp_o$ and
    \item $\left[\left(\Delta_1; A_{1}\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right]$ is a list of pairs where
    \item the $i$-th pair $(\Delta_i, A_i) :\Cxt_{\Sigma}(\Xi) \times \Term_{\Sigma}(\Xi)$ are types of the binding variables and the type of the $i$-th argument of $\termOp_{o}$.
  \end{enumerate}
  For brevity, we write $o \colon \Xi \vdash (\Delta_1)A_{1}, \ldots, \left[\Delta_{n}\right] A_{n} \to A$ to indicate an operation $o$ with its arity. 
\end{definition}
\begin{remark}[Terminology on binding signature]
  \cite{Aczel1978,Fiore2010}
\end{remark}

@@

Intrinsically typed terms for a type signature $\Sigma$ and a term signature $\Omega$ are nothing more than derivations of the intrinsic typing judgement $\boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A}$ constructed with only the rule $(\var)$ and the rule scheme $(\termOp)$ displayed in \Cref{fig:intrinsic-typing}.

The substitution $\rho\colon \Sub{\Xi}{\Theta}$ is used to instantiate variables in the local context $\Xi$ with types within the context of type variables in $\Theta$.
Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\termOp$ to ensure all types are well-formed under $\Theta$.

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A} \quad \text{where $\Theta \vdash_{\Sigma} \Gamma$ and $\Theta \vdash_{\Sigma} A$} \\
    \inferrule{A \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A}\;(\var)
    \and
    \inferrule{\rho : \Sub{\Xi}{\Theta} \\ \Theta \mid \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \quad\cdots\quad \Theta \mid \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
    {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\termOp)
    \and \text{for $o\colon \Xi \vdash [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
  \end{mathpar}
  \caption{Intrinsic typing rules}
  \label{fig:intrinsic-typing}
\end{figure}
\subsection{Raw Terms and Extrinsically Typed Terms?}
\begin{LTannote}{no extrinsic typing}
Argue that we don't need extrinsically typed terms but raw terms only as the output of some parser.
\end{LTannote}

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta \vdash_{\Sigma, \Omega} t}
  \end{mathpar}

  \caption{Raw terms}
\end{figure}

@@

\begin{definition}
  For a type signature $\Sigma$, a \emph{bidirectional binding signature} $\Omega$ is a set $O$ with a function
  \[
    \mathit{ar}\colon O \to \sum_{\Xi : \UU} \left(\Cxt_{\Sigma}(\Xi) \times \Mode \times \Type_{\Sigma}(\Xi)\right)^* \times \Mode \times \Type_{\Sigma}(\Xi).
  \]
  where $\Mode$ consists of $\chk$ for checking and $\syn$ for synthesis.
  Bidirectional binding signatures are just binding signatures (\Cref{def:binding-signature}) augmented with a mode for each argument and its target of a construct $\termOp_o$ in a bidirectional system, i.e.
  an arity is a $4$-tuple
  \[
    \left(\Xi, \left[\left(\Delta_1, d_1, A_1\right), \ldots, \left(\Delta_{n}, d_n, A_{n}\right) \right], d, A\right)
  \]
  where $d$ and $d_i$'s indicate the modes of a construct and its arguments respectively.

@@

\subsubsection{Bidirectional Raw Terms}

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta \vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}}
  \end{mathpar}

  \caption{Bidirectional raw terms}
\end{figure}

@@

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{d}}
    \quad \text{where $d$ is either $\chk$ or $\syn$, $\Theta \vdash_{\Sigma} \Gamma$, $\Theta \vdash_{\Sigma} A$, and $\Theta \vdash_{\Sigma, \Omega} t^d$} 
    \\
    \inferrule{(x : A) \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A^\syn}\;(\var)
    \and
    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^\chk \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t : B}:  A^\syn}\;(\text{anno})
    \and
    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^{\syn} \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t\uparrow} : A^\chk}\;(\text{sub})
  \end{mathpar}
  \begin{mathpar}
    \inferrule{\rho\colon \Sub{\Xi}{\Theta} \\
      \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} : \simsub{A_{1}}{\rho}^{\isDir{d_1}} \\
      \cdots \\
    \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}^{\isDir{d_n}}}
  {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{\termOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} : \simsub{A}{\rho}^{\isDir{d}}} \; (\termOp)
    \and \text{for $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
    \\
    \text{Abbreviations: $\boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A} \defeq \boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\syn}$ and $\boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} \defeq \boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\chk}$}
  \end{mathpar}
  \caption{Bidirectional extrinsic typing rules}
\end{figure}

>>>>

\usepackage{bm,braket,mathtools,mathpartir,bussproofs}
\EnableBpAbbreviations
\usepackage[all,cmtip]{xy}

\usepackage{xfrac,xspace,xcolor,subcaption}
\usepackage{hyperref}
\usepackage[capitalise,noabbrev]{cleveref}
\usepackage[inline]{enumitem} % for environment enumerate*

@@

\section{Bidirectional simply typed \texorpdfstring{$\lambda$}{lambda}-calculus}\label{sec:bi-stlc}
\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta \mid \Gamma \vdash \isTerm{t} \Rightarrow A}
    \quad\text{and}\quad \boxed{\Theta \mid \Gamma \vdash \isTerm{t} \Leftarrow A}
    \quad\text{where $\Theta \vdash \Gamma$ and $\Theta \vdash A$} 
    \\
    \inferrule{(x : A) \in \Gamma}{\Theta \mid \Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
    \and
    \inferrule{\Theta \mid \Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Theta \mid \Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
    \and
    \inferrule{\Theta \mid \Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash \isTerm{t\uparrow} \Leftarrow A}\;(\text{sub})
    \and
    \inferrule{\Theta \mid \Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Theta \mid \Gamma \vdash \isTerm{u} \Leftarrow A}{\Theta \mid \Gamma \vdash \isTerm{t\;u} \Rightarrow B}
    \and
    \inferrule{\Theta \mid \Gamma, \isTerm{x} : A \vdash \isTerm{t} \Leftarrow B}{\Theta \mid \Gamma \vdash \isTerm{\lam{x}t} \Leftarrow A \bto B}
  \end{mathpar}
  \caption{The extrinsic style of bidirectional simply typed $\lambda$-calculus with type variables}
  \label{fig:bi-stlc}
\end{figure}
\begin{remark}
  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
\end{remark}

@@

\subsection{Binding Signatures}
\begin{definition}\label{def:binding-signature}
  For a type signature $\Sigma$, a \emph{binding signature} $\Omega$ consists of a set $O$ and a function
  \[
    \mathit{ar}\colon O \to \sum_{\Xi : \UU} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
  \]
  Each inhabitant $o: O$ is meant to represent a term construct $\tmOp_o$ in a simple type theory with a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A\right)$
  instead of a number as its arity $\arity_o$ where
  \begin{enumerate}
    \item $A : \Type_\Sigma(\Xi)$ is the target type of $\tmOp_o$ and
    \item $\left[\left(\Delta_1; A_{1}\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right]$ is a list of pairs where
    \item the $i$-th pair $(\Delta_i, A_i) :\Cxt_{\Sigma}(\Xi) \times \Term_{\Sigma}(\Xi)$ are types of the binding variables and the type of the $i$-th argument of $\tmOp_{o}$.
  \end{enumerate}
  For brevity, we write $o \colon \Xi \vdash (\Delta_1)A_{1}, \ldots, \left[\Delta_{n}\right] A_{n} \to A$ to indicate an operation $o$ with its arity. 
\end{definition}
\begin{remark}[Terminology on binding signature]
  \cite{Aczel1978,Fiore2010}
\end{remark}

@@

Intrinsically typed terms for a type signature $\Sigma$ and a term signature $\Omega$ are nothing more than derivations of the intrinsic typing judgement $\boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A}$ constructed with only the rule $(\var)$ and the rule scheme $(\tmOp)$ displayed in \Cref{fig:intrinsic-typing}.

The substitution $\rho\colon \Sub{\Xi}{\Theta}$ is used to instantiate variables in the local context $\Xi$ with types within the context of type variables in $\Theta$.
Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\tmOp$ to ensure all types are well-formed under $\Theta$.

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A} \quad \text{where $\Theta \vdash_{\Sigma} \Gamma$ and $\Theta \vdash_{\Sigma} A$} \\
    \inferrule{A \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A}\;(\var)
    \and
    \inferrule{\Theta \mid \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \quad\cdots\quad \Theta \mid \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
    {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\tmOp)
    \and \text{for $o\colon \Xi \vdash [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$
    and $\rho : \Sub{\Xi}{\Theta}$}
  \end{mathpar}
  \caption{Intrinsic typing rules}
  \label{fig:intrinsic-typing}
\end{figure}

\subsection{Raw Terms and Extrinsically Typed Terms}

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta \vdash_{\Sigma, \Omega} \isTerm{t}}
    \\
    \inferrule{x : \Identifier}{\Theta \vdash_{\Sigma, \Omega} \isTerm{x}}\;(\text{var})
    \and
    \inferrule{\Theta \vdash A \\ \Theta \vdash_{\Sigma, \Omega}\isTerm{t}}{\Theta \vdash_{\Sigma, \Theta} \isTerm{t \annote A}}\;(\text{anno})
    \\
    \inferrule{\Theta \vdash_{\Sigma, \Omega} \isTerm{t_1} \quad \cdots \quad \Theta \vdash_{\Sigma, \Omega} \isTerm{t_n}}
    {\Theta \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(t_1, \ldots, t_n)}}\;(\text{op}) 
    \and \text{for $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
  \end{mathpar}
  \caption{Annotatable raw terms}
\end{figure}

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A} \quad \text{where $\Theta \vdash_{\Sigma} \Gamma$, $\Theta \vdash_{\Sigma} A$, and $\Theta \vdash_{\Sigma, \Omega} t$} \\
    \inferrule{(x : A) \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A}\;(\var)
    \and
    \inferrule{\Theta \mid \Gamma \vdash \isTerm{t} : A}{\Theta \mid \Gamma \vdash (\isTerm{t \annote A}) : A}\;(\text{anno})
    \and
    \inferrule{\Theta \mid \Gamma, \isTerm{\vec{x}_1} : \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Theta \mid \Gamma, \isTerm{\vec{x}_n} : \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
    {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\,t_1; \ldots; \vec{x}_n.\,t_n)} : \simsub{A}{\rho}}\;(\tmOp)
    \and \text{for $o\colon \Xi \vdash [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$
    and $\rho : \Sub{\Xi}{\Theta}$}
  \end{mathpar}
  \caption{Extrinsic typing rules}
  \label{fig:extrinsic-typing}
\end{figure}

@@

\begin{definition}
  For a type signature $\Sigma$, a \emph{bidirectional binding signature} $\Omega$ is a set $O$ with a function
  \[
    \mathit{ar}\colon O \to \sum_{\Xi : \UU} \left(\Cxt_{\Sigma}(\Xi) \times \Mode \times \Type_{\Sigma}(\Xi)\right)^* \times \Mode \times \Type_{\Sigma}(\Xi).
  \]
  where $\Mode$ consists of $\chk$ for checking and $\syn$ for synthesis.
  Bidirectional binding signatures are just binding signatures (\Cref{def:binding-signature}) augmented with a mode for each argument and its target of a construct $\tmOp_o$ in a bidirectional system, i.e.
  an arity is a $4$-tuple
  \[
    \left(\Xi, \left[\left(\Delta_1, d_1, A_1\right), \ldots, \left(\Delta_{n}, d_n, A_{n}\right) \right], d, A\right)
  \]
  where $d$ and $d_i$'s indicate the modes of a construct and its arguments respectively.

@@

@@

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta \vdash_{\Sigma, \Omega} \isTerm{t}\;\isDir{d}}
    \\
    \inferrule{x : \Identifier}{\Theta \vdash_{\Sigma, \Omega} \isTerm{x}\;\isDir{\syn}}\;(\text{var})
    \and
    \inferrule{\Theta \vdash A \\ \Theta \vdash_{\Sigma, \Omega}\isTerm{t}\;\isDir{\chk}}{\Theta \vdash_{\Sigma, \Theta} \isTerm{t \annote A}\;\isDir{\syn}}\;(\text{anno})
    \and
%    \inferrule{\Theta \vdash_{\Sigma, \Omega}\isTerm{t}\;\isDir{\chk}}{\Theta, X \vdash_{\Sigma, \Theta} \isTerm{t \downarrow}\;\isDir{\syn}}\;(\text{anno}')
%    \and
    \inferrule{\Theta \vdash_{\Sigma, \Omega} \isTerm{t}\;\isDir{\syn}}{\Theta \vdash_{\Sigma, \Omega} (\isTerm{t \uparrow})\;\isDir{\chk}}
  \end{mathpar}
  \begin{mathpar}
    \inferrule{\rho \colon \Sub{\Xi}{\Theta} \\ \Theta \vdash_{\Sigma, \Omega} \isTerm{t_1}\;\isDir{d_1} \quad \cdots \quad \Theta \vdash_{\Sigma, \Omega} \isTerm{t_n}\;\isDir{d_n}}
    {\Theta \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\rho; \vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}\;\isDir{d}}\;(\text{op})
    \and \text{for $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
  \end{mathpar}

  \caption{Bidirectional pre-typed terms}
\end{figure}
\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{d}}
    \quad \text{where $d$ is either $\chk$ or $\syn$, $\Theta \vdash_{\Sigma} \Gamma$, $\Theta \vdash_{\Sigma} A$, and $\Theta \vdash_{\Sigma, \Omega} t\;d$} 
    \\
    \inferrule{(x : A) \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A^\syn}\;(\var)
    \and
    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^\chk \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} (\isTerm{t \annote B}):  A^\syn}\;(\text{anno})
    \and
    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^{\syn} \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t\uparrow} : A^\chk}\;(\text{sub})
    \\
    \inferrule{\rho\colon \Sub{\Xi}{\Theta} \\ \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} : \simsub{A_{1}}{\rho}^{\isDir{d_1}} \\
      \cdots \\
    \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}^{\isDir{d_n}}}
  {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\rho; \vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} : \simsub{A}{\rho}^{\isDir{d}}} \; (\tmOp)
    \and \text{for $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
    \\
    \text{Abbreviations: $\boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A} \defeq \boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\syn}$ and $\boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} \defeq \boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\chk}$}
  \end{mathpar}
  \caption{Bidirectional extrinsic typing rules}
\end{figure}
